Exploring the role of electric field in the catalase active site

### 2018-03-16 bb494/manuscript.tex
all

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Start the main part of the manuscript here.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Catalase is an enzyme which catalyzes the breakdown of hydrogen peroxide into water and oxygen gas. This is an important detoxification reaction, since hydrogen peroxide is a reactive toxin generated by several biological processes. Catalase is a tetrameric protein which uses a heme group in its active site to facilitate the breakdown of hydrogen peroxide. The molecular mechaninism of catalase proceeds in two steps. First, the heme reduces a hydrogen peroxide, producing water and the so-called Cpd I. Then Cpd I oxidizes a second hydrogen peroxide, producing water and oxygen gas. This second step is mediated by a histidine near the active site \cite{alfonso-prieto}.

Catalase's activity is nearly diffusion-limited \cite{kcatkm, difflimit}. This rate suggests that nearly every step in the substrate-enzyme interaction is optimized for speed. Understanding the optimizations of diffusion-limited proteins can provide many insights into how proteins achieve their functions. Catalase has a variety of interesting quirks for funneling substrate into its active pocket. The enzyme has a number of high-H2O2-residency surface residues near its active channel, which are believed to concentrate the substrate near the entrance to the active channel \cite{concentrateh2o2}. The active channel itself seems optimized to maximize substrate flow, to the point that mutagenesis to open the channel decreases enzyme activity \cite{substrateflow}. The channel seems to be capable of selecting the molecules which enter it using a patch of hydrophobic residues which are situated to preferentially pass the bulkier H2O2 \cite{molecularruler}. Even the directionality of substrate within the channel seems to be controlled: one channel is used for substrate entrance, and the other for product exit \cite{lateralchannel}. An new finding suggests that catalase might control not just the presence of hydrogen peroxide, but even its orientation as it enters the active site. It is believed that an electric field from a negative Asn residue to the positive heme acts to orient the H2O2 dipole, positioning the molecule such that it enters the active site at an angle ideal for reacting \cite{electricpotential}. This use of an electric field over a long range to create an ``orienting field'' is a new and exciting usage of residues to achieve a function; understanding its role in catalase may help to understand the mechanisms of other proteins.

In this study we hope to study the proposed orienting field optimization and its influence on catalase activity. In particular, we hope to carefully disrupt the wildtype active site potential environment by introducing a new charged residue, then examine the effect this disruption has on activity. For meaningful results, we must be careful to not introduce structural modifications to the enzyme. To ensure this, we will use genetic code expansion to incorporate 3-nitrotyrosine in place of tyrosine 206. The T206NY mutation will add only 30 $\AA^3$ of steric bulk \cite{3ntsize}, so should have minimal influence on active site structure. The theorized results of this addition are diagrammed in Figure \ref{fig:hypothesis}. To study the influence of field disruption on activity, we will study the activity of T206NY above and below its pKa of 6.8 \cite{3ntsize}. Wildtype catalase activity is constant under pH values from 5 to 10 \cite{phdependence,kcatkm}, so pH-dependence should only be seen if electric environment in the active site is important for activity.\\

\begin{figure}
  \begin{subfigure}{0.69\textwidth}
    \begin{minipage}{0.1\textwidth}\caption{}\end{minipage}%
    \begin{minipage}{.9\textwidth}
      \includegraphics[width=0.5\linewidth]{figures/cartoon-visualization.png}%
      \includegraphics[width=0.5\linewidth]{figures/channel-visualization.png}
    \end{minipage}
  \end{subfigure}
  \begin{subfigure}{0.3\textwidth}
    \begin{minipage}{0.1\textwidth}\caption{}\end{minipage}%
    \begin{minipage}{.9\textwidth}
      \includegraphics[width=\linewidth]{figures/active-site-visualization.png}%
    \end{minipage}
  \end{subfigure}

  \caption{Depiction of catalase active pocket. \textbf{Left:} wildtype pocket with proposed ligand-orienting E-field \cite{electricpotential}. The field should orient H2O2 for optimal reaction with the heme. \textbf{Right:} modified HPII-3NY with a deprotonated 3-nitrotyrosine residue at site 206, with hypothetical augmented E-field. The disrupted E-field should result in un- or mis-aligned substrate.}
  \label{fig:hypothesis}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Start the main part of the manuscript here.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%% Catalase is an enzyme which catalyzes the breakdown of hydrogen peroxide into water and oxygen gas. This is an important detoxification reaction, since hydrogen peroxide is a reactive toxin generated by several biological processes. Catalase is a tetrameric protein which uses a heme group in its active site to facilitate the breakdown of hydrogen peroxide. The molecular mechaninism of catalase proceeds in two steps. First, the heme reduces a hydrogen peroxide, producing water and the so-called Cpd I. Then Cpd I oxidizes a second hydrogen peroxide, producing water and oxygen gas. This second step is mediated by a histidine near the active site \cite{alfonso-prieto}.

%% Catalase's activity is nearly diffusion-limited \cite{kcatkm, difflimit}. This rate suggests that nearly every step in the substrate-enzyme interaction is optimized for speed. Understanding the optimizations of diffusion-limited proteins can provide many insights into how proteins achieve their functions. Catalase has a variety of interesting quirks for funneling substrate into its active pocket. The enzyme has a number of high-H2O2-residency surface residues near its active channel, which are believed to concentrate the substrate near the entrance to the active channel \cite{concentrateh2o2}. The active channel itself seems optimized to maximize substrate flow, to the point that mutagenesis to open the channel decreases enzyme activity \cite{substrateflow}. The channel seems to be capable of selecting the molecules which enter it using a patch of hydrophobic residues which are situated to preferentially pass the bulkier H2O2 \cite{molecularruler}. Even the directionality of substrate within the channel seems to be controlled: one channel is used for substrate entrance, and the other for product exit \cite{lateralchannel}. An new finding suggests that catalase might control not just the presence of hydrogen peroxide, but even its orientation as it enters the active site. It is believed that an electric field from a negative Asn residue to the positive heme acts to orient the H2O2 dipole, positioning the molecule such that it enters the active site at an angle ideal for reacting \cite{electricpotential}. This use of an electric field over a long range to create an ``orienting field'' is a new and exciting usage of residues to achieve a function; understanding its role in catalase may help to understand the mechanisms of other proteins.

%% In this study we hope to study the proposed orienting field optimization and its influence on catalase activity. In particular, we hope to carefully disrupt the wildtype active site potential environment by introducing a new charged residue, then examine the effect this disruption has on activity. For meaningful results, we must be careful to not introduce structural modifications to the enzyme. To ensure this, we will use genetic code expansion to incorporate 3-nitrotyrosine in place of tyrosine 206. The T206NY mutation will add only 30 $\AA^3$ of steric bulk \cite{3ntsize}, so should have minimal influence on active site structure. The theorized results of this addition are diagrammed in Figure \ref{fig:hypothesis}. To study the influence of field disruption on activity, we will study the activity of T206NY above and below its pKa of 6.8 \cite{3ntsize}. Wildtype catalase activity is constant under pH values from 5 to 10 \cite{phdependence,kcatkm}, so pH-dependence should only be seen if electric environment in the active site is important for activity.\\

Catalase is an enzyme which catalyzes the breakdown of hydrogen peroxide into water and oxygen gas. This is an important detoxification reaction, since hydrogen peroxide is a reactive toxin generated by several biological processes. Catalase is a tetrameric protein which uses a heme group in its active site to facilitate the breakdown of hydrogen peroxide. The molecular mechaninism of catalase proceeds in two steps. First, the heme reduces a hydrogen peroxide, producing water and the so-called Cpd I. Then Cpd I oxidizes a second hydrogen peroxide, producing water and oxygen gas. This second step is mediated by a histidine near the active site \cite{alfonso-prieto}.\\

One remaining question about catalase is how it achieves its remarkable speed given its buried active site. The enzyme is nearly diffusion-limited\cite{kcatkm, difflimit}, indicating it turns substrate to product at close to the optimal rate. However, in order to reach the active site, hydrogen peroxide must diffuse through a 30 $\AA$ tunnel\cite{substrateflow}. Thus, substrate must be transported very efficiently through the protein. Understanding how this transport process works could provide lots of insight into how enzymes traffic substrate into buried active sites. Catalase is one of the most studied proteins in existence, and so represents a good model protein for understanding substrate trafficking.\\

Previous work on substrate trafficking has illuminated various stages in the process. The enzyme has a number of high-H2O2-residency surface residues, which are believed to concentrate the substrate near the entrance to the active channel \cite{concentrateh2o2}. The active channel itself seems optimized to maximize substrate flow, to the point that mutagenesis to open the channel decreases enzyme activity \cite{substrateflow}. The channel seems to be capable of selecting the molecules which enter it using a patch of hydrophobic residues which are situated to preferentially pass the bulkier H2O2 \cite{molecularruler}. Even the directionality of substrate within the channel seems to be controlled: one channel is used for substrate entrance, and the other for product exit \cite{lateralchannel}. However, how substrate is traficked once in the active site antechamber is still an open question.\\

One tool proteins are known to use to speed up active site kinetics is the use of electric fields. Superoxide dismutase is known to have a strong E-field in its active site to stabilize dipolar intermediates \cite{conserved-as-efield-sod, concentrated-as-efield-sod}. Several other proteins are known to use a similar effect \cite{efield-review} (be more precise here). It has been suggested that catalase uses an electric field in its active site to orient and traffick hydrogen peroxide to its active heme \cite{electricpotential}. This model, proposed by Chelikani \textit{et. al}, theorizes that a strong electric field from the negative asparigine 181 at the top of the active site antechamber extends to the strongly positive heme iron, as shown in Figure \ref{fig:hypothesis}.d. This electric field would orient incoming hydrogen peroxides in the optimal rotation for reaction with heme.\\

An issue with mutagenesis studies like Chelikani \textit{et. al.} is that catalase's active site is very sensitive to changes. Mutagenesis of many residues results in large changes in enzyme kinetics \cite{substrateflow}. Here we conduct a less invasive experiment to probe the role of electric field in the catalase active site. We use genetic code expansion \cite{hammill} to incorporate two functional groups onto F206 in the catalase active site, turning it into 3-nitrotyrosine (Figure \ref{fig:hypothesis}.e). This change is important because it a.) is tiny, adding only 30 $\AA^3$ of steric bulk \cite{3ntsize}, and b.) incorporates a hydroxy group of pKa 7.0 into the catalase active site \cite{3ntsize}. Catalase activity is pH-invariant in the range of 5 to 9 \cite{phdependence,kcatkm}; thus, by incorporating a group which gains charge at pH 7, we can precisely study the effect of charge in the catalase active site on enzyme kinetics without worrying about the undesired effects of mutagenesis.\\

\begin{figure}
  \begin{subfigure}{0.49\textwidth}
    \begin{minipage}{0.1\textwidth}\caption{}\end{minipage}%
    \begin{minipage}{.9\textwidth}
      %% \includegraphics[width=0.5\linewidth]{figures/cartoon-visualization.png}%
      \includegraphics[width=\linewidth]{figures/channel-visualization.png}
    \end{minipage}
  \end{subfigure}
  \begin{subfigure}{0.5\textwidth}
    \begin{minipage}{0.1\textwidth}\caption{}\end{minipage}%
    \begin{minipage}{.9\textwidth}
      \includegraphics[width=0.5\linewidth]{figures/active-site-visualization.png}%
      \includegraphics[width=0.5\linewidth]{figures/active-site-visualization.png}%
    \end{minipage}
  \end{subfigure}
  \caption{Depiction of catalase active pocket. \textbf{Left:} wildtype pocket with proposed ligand-orienting E-field \cite{electricpotential}. The field should orient H2O2 for optimal reaction with the heme. \textbf{Right:} modified HPII-3NY with a deprotonated 3-nitrotyrosine residue at site 206, with hypothetical augmented E-field. The disrupted E-field should result in un- or mis-aligned substrate.}
  \label{fig:hypothesis}
\end{figure}
